
---
title: " Hybrid Forecasting (Prophet + XGBoost) of Philippine Leptospirosis Cases (2019–2023)"
Project Overview: This project implements a Hybrid Residual Learning Model (Prophet + XGBoost) to forecast weekly Leptospirosis cases.

Prophet models the global trend and seasonality (baseline).

XGBoost models the residuals (errors) of Prophet to correct for non-linear outbreaks and intensity misses.

Data Source: [Philippine Epidemiological Data](https://data.mendeley.com/datasets/v98325dpxv/1)

---


### 1. Setup & Dependencies
Loads necessary libraries for time series analysis, machine learning, and visualization.

```{r}
# --- 1. SETUP & DEPENDENCIES ---
library(tidyverse)  # Data manipulation & plotting
library(zoo)        # Interpolation (na.approx)
library(lubridate)  # Date handling
library(prophet)    # Baseline time series forecasting
library(xgboost)    # Gradient boosting (Residual correction)
library(Metrics)    # Accuracy metrics (MAE, RMSE)
library(Matrix)     # Sparse matrices for XGBoost
library(patchwork)  # Combining plots (for better paper visuals)
library(forecast)   # for SARIMA

# Set global theme for all plots
theme_set(theme_minimal())
```


### 2. Data Loading & Processing
Reads the raw CSV, parses dates, and ensures a continuous weekly timeline (filling missing weeks with NA).
```{r}
# --- 2. LOAD & PROCESS DATES ---
raw_data <- read_csv("Leptospirosis_WeeklyCasesPh_2019-2023.csv")

clean_data <- raw_data %>%
  mutate(
    Date_Parsed = dmy(Date), 
    Year = year(Date_Parsed), 
    Week = week(Date_Parsed)
  ) %>%
  select(Year, Week, Cases, Date_Parsed) %>%
  mutate(Cases = as.numeric(Cases)) %>%
  
  # Ensure every week (1-52) exists for every year (2019-2023)
  complete(Year = 2019:2023, Week = 1:52, fill = list(Cases = NA)) %>%
  
  # Reconstruct Dates for the newly created empty rows
  mutate(
    Date_Parsed = if_else(
      is.na(Date_Parsed),
      ymd(paste0(Year, "-01-01")) + weeks(Week - 1),
      Date_Parsed
    )
  ) %>%
  
  # Force Specific 2020 Gaps to NA (Lockdown/Typhoon Logic for Imputation)
  mutate(Cases = replace(Cases, Year == 2020 & Week >= 11 & Week <= 26, NA)) %>%
  mutate(Cases = replace(Cases, Year == 2020 & Week >= 35 & Week <= 38, NA)) %>%
  mutate(Cases = replace(Cases, Year == 2020 & Week >= 43 & Week <= 45, NA))

print(paste("Total Missing Dates:", sum(is.na(clean_data$Date_Parsed))))
```

### 3. Context-Aware Imputation
Fills missing values using a layered strategy: Seasonal Average + Event Adjustment (Lockdown/Typhoon) + Linear Interpolation.

```{r}
# --- 3. CONTEXT-AWARE IMPUTATION ---

# Step A: Calculate "Seasonal Baseline" (Excluding 2020 anomalies)
seasonal_profile <- clean_data %>%
  filter(Year != 2020) %>%
  group_by(Week) %>%
  summarise(Seasonal_Avg = mean(Cases, na.rm = TRUE), .groups = "drop")

# Step B: Apply Imputation Logic
imputed_data <- clean_data %>%
  left_join(seasonal_profile, by = "Week") %>%
  mutate(
    Type = ifelse(is.na(Cases), "Imputed", "Observed"),
    
    Cases_Imputed = case_when(
      # 1. COVID LOCKDOWN (2020) -> 50% of Normal Seasonal Baseline
      Year == 2020 & Week >= 11 & Week <= 26 & is.na(Cases) ~ Seasonal_Avg * 0.5,
      
      # 2. TYPHOON SEASON (2020) -> 100% of Normal Seasonal Baseline
      Year == 2020 & Week >= 43 & Week <= 45 & is.na(Cases) ~ Seasonal_Avg,
      
      # 3. Keep Observed Data
      !is.na(Cases) ~ Cases,
      
      # 4. Remaining gaps -> NA (to be filled by linear interpolation)
      TRUE ~ NA_real_
    )
  ) %>%
  # Linear Interpolation for tiny gaps
  mutate(Cases_Imputed = na.approx(Cases_Imputed, na.rm = FALSE))

# Final Prep for Modeling: Remove future empty rows (Nov-Dec 2023)
model_data <- imputed_data %>%
  select(ds = Date_Parsed, y = Cases_Imputed, Week, Year) %>%
  filter(!is.na(y)) %>%  # Essential fix to prevent validation crashes
  arrange(ds)

print(paste("Data Ready. Total rows:", nrow(model_data)))
```

### 4. Model Configuration
Sets training windows, XGBoost hyperparameters, and date indices for the rolling validation loop.

```{r}
# --- 4. CONFIGURATION ---

# Training Parameters
initial_train_weeks <- 115   # Starting window size
prediction_horizon  <- 2     # Predict 2 weeks ahead
slide_step          <- 16    # Test every ~4 months

# XGBoost Parameters (Corrective Model)
xgb_params <- list(
  objective = "reg:squarederror",
  max_depth = 1,        # Decision stumps (prevents overfitting on small data)
  learning_rate = 0.35  # Aggressive correction
)

# Date Indexing
unique_dates <- sort(unique(model_data$ds))
start_index <- initial_train_weeks
end_index   <- length(unique_dates) - prediction_horizon
prophet_fold_data <- list() # Storage for cross-validation folds

print(paste("Validation Range Indices:", start_index, "to", end_index))
```

### 5. Phase 1: Prophet Baseline & Residual Extraction
Trains Prophet on rolling windows, generates forecasts, and calculates "Residuals" (errors) for XGBoost to learn.

```{r}
# --- 5. PHASE 1: PROPHET BASELINE ---
print("Generating Baselines...")

for (i in seq(start_index, end_index, by = slide_step)) {
  
  # Define Window
  cutoff_date <- unique_dates[i]
  cv_train <- model_data %>% filter(ds <= cutoff_date)
  cv_test  <- model_data %>% filter(ds > cutoff_date & ds <= unique_dates[i + prediction_horizon])
  
  if(nrow(cv_test) < prediction_horizon) break 
  
  # Train Prophet
  m <- prophet(cv_train, 
               daily.seasonality = FALSE,
               weekly.seasonality = TRUE, 
               yearly.seasonality = TRUE,
               seasonality.mode = 'multiplicative',
               changepoint.prior.scale = 0.55, #best 0.55
               seasonality.prior.scale = 5)           
  
  # Generate Forecasts
  forecast_train <- predict(m, m$history) %>% select(ds = ds, yhat) %>% mutate(ds = as.Date(ds))
  
  future_test  <- make_future_dataframe(m, periods = nrow(cv_test), freq = "week")
  forecast_test <- predict(m, future_test) %>% select(ds = ds, yhat) %>% mutate(ds = as.Date(ds)) %>%
    filter(ds %in% cv_test$ds)
  
  # Calculate Residuals (Training Data)
  # Residual = Actual - Predicted. This is what XGBoost will try to predict.
  cv_train_prepared <- cv_train %>%
    inner_join(forecast_train, by = "ds") %>%
    mutate(Residual = y - yhat, WeekNum = isoweek(ds))
  
  # Prepare Test Data
  cv_test_prepared <- cv_test %>%
    inner_join(forecast_test, by = "ds") %>%
    mutate(WeekNum = isoweek(ds))

  # Store Fold Data
  fold_id <- paste0("fold_", i)
  prophet_fold_data[[fold_id]] <- list(train = cv_train_prepared, test = cv_test_prepared)
  cat(".") 
}
print("Phase 1 Complete.")
```


### 6. Phase 2: Benchmarking (SARIMA + Seasonal Naive)
This generates the standard baselines to compare against the Hybrid mode
```{r}
# --- 6. PHASE 2: BENCHMARKS (SARIMA & NAIVE) ---
print("Generating Benchmarks (SARIMA & Seasonal Naive)...")

benchmark_results_list <- list()

for (fold_id in names(prophet_fold_data)) {
  
  # 1. Retrieve Data from Phase 1
  fold_data <- prophet_fold_data[[fold_id]]
  cv_train  <- fold_data$train
  cv_test   <- fold_data$test
  
  # --- MODEL A: SEASONAL NAIVE (The "Floor") ---
  # Prediction: Cases next week = Cases exactly 52 weeks ago
  fold_snaive <- cv_test %>%
    rowwise() %>%
    mutate(
      target_date_last_year = ds - weeks(52),
      # Lookup historical value from the main clean dataset
      SNaive_Forecast = model_data$y[model_data$ds == target_date_last_year][1]
    ) %>%
    ungroup() %>%
    # Fill gaps (rare) with recent mean if last year's data is missing
    mutate(SNaive_Forecast = ifelse(is.na(SNaive_Forecast), mean(cv_train$y), SNaive_Forecast))

  # --- MODEL B: SARIMA (The "Standard") ---
  # Create Time Series Object (Freq=52 for weekly data)
  ts_train <- ts(cv_train$y, frequency = 52)
  
  # Train SARIMA (Stepwise=FALSE is slower but more accurate for publication)
  fit_sarima <- auto.arima(ts_train, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
  fcst_sarima <- forecast(fit_sarima, h = nrow(cv_test))
  
  # Combine Benchmark Results
  fold_benchmarks <- fold_snaive %>%
    mutate(
      SARIMA_Forecast = as.numeric(fcst_sarima$mean),
      SARIMA_Forecast = pmax(SARIMA_Forecast, 0) # Remove negatives
    ) %>%
    select(ds, SNaive_Forecast, SARIMA_Forecast)
  
  benchmark_results_list[[fold_id]] <- fold_benchmarks
  cat(".")
}

benchmark_results_df <- bind_rows(benchmark_results_list)
print("Benchmarking Complete.")
```







### 7. Phase 3: Hybrid Calibration & Merging
 It runs XGBoost and then merges all model results (Prophet, XGBoost, SARIMA, SNaive) into one master table.

```{r}
# --- 7. PHASE 3: XGBOOST & MERGING ---
print("Training XGBoost Corrections...")

hybrid_results_list <- list()
set.seed(123)

for (fold_id in names(prophet_fold_data)) {

  fold_data <- prophet_fold_data[[fold_id]]
  cv_train <- fold_data$train
  cv_test  <- fold_data$test

  # XGBoost Matrix Creation
  dtrain <- sparse.model.matrix(~ WeekNum + yhat, data = cv_train)
  dtest  <- sparse.model.matrix(~ WeekNum + yhat, data = cv_test)
  dtrain_xgb <- xgb.DMatrix(data = dtrain, label = cv_train$Residual)
  dtest_xgb  <- xgb.DMatrix(data = dtest)

  # Train & Predict
  xgb_cv <- xgb.train(
    params = list(objective = "reg:squarederror", max_depth = 1, eta = 0.35),
    data = dtrain_xgb, nrounds = 500, verbose = 0
  )
  correction <- predict(xgb_cv, dtest_xgb)

  # Combine Results
  fold_result <- cv_test %>%
    mutate(
      Prophet_Only = yhat,
      Hybrid_Forecast = pmax(yhat + correction, 0)
    )
  
  hybrid_results_list[[fold_id]] <- fold_result
}

# --- MASTER MERGE ---
# Join Hybrid results with Benchmark results
rolling_results <- bind_rows(hybrid_results_list) %>%
  left_join(benchmark_results_df, by = "ds") %>%
  distinct(ds, .keep_all = TRUE)

print("All Models Merged.")
```


### 8. Evaluation & Metrics
This generates the "Scorecard" table comparing all 4 models.

```{r}
# --- 8. PUBLICATION EVALUATION ---

# 1. Define Custom Metric Functions
# SMAPE: Symmetric Mean Absolute Percentage Error (Good for seasonal data)
smape_custom <- function(actual, pred) {
  # Avoid division by zero by adding a tiny epsilon
  mean(2 * abs(pred - actual) / (abs(actual) + abs(pred) + 1e-6)) * 100 
}

# Bias: Mean Error (Negative = Under-prediction, Positive = Over-prediction)
bias_custom <- function(actual, pred) {
  mean(pred - actual)
}

# 2. Scorecard Generator
print_scorecard <- function(df) {
  df %>%
    summarise(
      # --- MAE (Precision) ---
      MAE_SNaive   = mae(y, SNaive_Forecast),
      MAE_SARIMA   = mae(y, SARIMA_Forecast),
      MAE_Prophet  = mae(y, Prophet_Only),
      MAE_Hybrid   = mae(y, Hybrid_Forecast),
      
      # --- RMSE (Outbreak Sensitivity) ---
      RMSE_SNaive  = rmse(y, SNaive_Forecast),
      RMSE_SARIMA  = rmse(y, SARIMA_Forecast),
      RMSE_Prophet = rmse(y, Prophet_Only),
      RMSE_Hybrid  = rmse(y, Hybrid_Forecast),
      
      # --- SMAPE (Relative Error %) ---
      SMAPE_SNaive   = smape_custom(y, SNaive_Forecast),
      SMAPE_SARIMA   = smape_custom(y, SARIMA_Forecast),
      SMAPE_Prophet  = smape_custom(y, Prophet_Only),
      SMAPE_Hybrid   = smape_custom(y, Hybrid_Forecast),

      # --- BIAS (Systematic Error) ---
      Bias_SNaive   = bias_custom(y, SNaive_Forecast),
      Bias_SARIMA   = bias_custom(y, SARIMA_Forecast),
      Bias_Prophet  = bias_custom(y, Prophet_Only),
      Bias_Hybrid   = bias_custom(y, Hybrid_Forecast)
    ) %>%
    pivot_longer(everything(), names_to = "Metric_Model", values_to = "Score") %>%
    # Split "MAE_Hybrid" into "MAE" and "Hybrid"
    separate(Metric_Model, into = c("Metric", "Model"), sep = "_") %>%
    pivot_wider(names_from = Metric, values_from = Score) %>%
    arrange(MAE) # Sort by best accuracy
}

# 3. Print Results
print("--- OVERALL MODEL RANKING (2019-2023) ---")
final_scorecard <- print_scorecard(rolling_results)
knitr::kable(final_scorecard, caption = "Overall Performance Metrics")

# Peak Analysis (High Cases Only)
threshold <- quantile(rolling_results$y, 0.90)  
print(paste("--- OUTBREAK DETECTION (Cases >", round(threshold,0), ") ---"))
peak_scorecard <- rolling_results %>% filter(y >= threshold) %>% print_scorecard()
knitr::kable(peak_scorecard, caption = "Performance During Outbreaks ( 90th percentile of cases)")
```


### 8. Comparative Visualization
This plots the top 3 contenders to keep the chart readable.

```{r}
# --- 9. VISUALIZATION (ALL MODELS) ---
plot_data <- rolling_results %>% distinct(ds, .keep_all = TRUE)


ggplot(plot_data, aes(x = ds)) +
  # 1. Actual Data (Grey Background)
  geom_line(aes(y = y, color = "Observed Cases"), linewidth = 1.2, alpha = 0.3) +
  
  # 2. Baselines (Dashed/Dotted)
  geom_line(aes(y = SNaive_Forecast, color = "SNaive"), linetype = "longdash", linewidth = 0.6) +
  geom_line(aes(y = Prophet_Only, color = "Prophet"), linetype = "longdash", linewidth = 0.6) +
  geom_line(aes(y = SARIMA_Forecast, color = "SARIMA"), linetype = "longdash", linewidth = 0.7) +
  
  # 3. Hybrid (The Proposed - Solid Red)
  geom_line(aes(y = Hybrid_Forecast, color = "Hybrid (Prophet+XGB)"), linewidth = 0.8) +
  
  # Styling
  scale_color_manual(values = c(
    "Observed Cases" = "gray20", 
    "SNaive" = "forestgreen",
    "Prophet" = "purple",
    "SARIMA" = "dodgerblue", 
    "Hybrid (Prophet+XGB)" = "firebrick"
  )) +
  labs(
    title = "Model Forecast Comparison: Hybrid vs. Baselines",
    y = "Weekly Leptospirosis Cases",
    x = "Date",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    plot.title = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

### 9. Interpretability (SHAP & Components)
Visualizes WHY the model works: Prophet's components + XGBoost's SHAP values.
```{r}
# --- 9. INTERPRETABILITY ---

# A. Prophet Components
# -----------------------
m_final <- prophet(model_data, weekly.seasonality=TRUE, yearly.seasonality=TRUE)
forecast_final <- predict(m_final, model_data)
prophet_plots <- prophet_plot_components(m_final, forecast_final)

# Combine using patchwork
final_prophet_plot <- (prophet_plots[[1]] + labs(title = "A. Global Trend")) /
                      (prophet_plots[[2]] + labs(title = "B. Weekly Seasonality")) /
                      (prophet_plots[[3]] + labs(title = "C. Yearly Seasonality")) +
                      plot_annotation(title = "Prophet Baseline Components")
print(final_prophet_plot)

# B. XGBoost SHAP Values (FINAL INTERPRETATION MODEL)
# --------------------------------------------------

# 1. Refit Prophet on FULL dataset (already used for interpretation)
m_final <- prophet(
  model_data,
  weekly.seasonality = TRUE,
  yearly.seasonality = TRUE,
  seasonality.mode = "multiplicative"
)

forecast_final <- predict(m_final, model_data)

# 2. Compute residuals on FULL dataset
residual_data <- model_data %>%
  left_join(
    forecast_final %>% select(ds, yhat),
    by = "ds"
  ) %>%
  mutate(
    Residual = y - yhat,
    WeekNum = isoweek(ds)
  )

# 3. Train ONE final XGBoost residual model (for SHAP only)
dmat_full <- sparse.model.matrix(~ WeekNum + yhat, data = residual_data)
dtrain_full <- xgb.DMatrix(data = dmat_full, label = residual_data$Residual)

xgb_final <- xgb.train(
  params = list(
    objective = "reg:squarederror",
    max_depth = 1,
    eta = 0.35
  ),
  data = dtrain_full,
  nrounds = 500,
  verbose = 0
)

# 4. Compute SHAP values
shap_contribs <- predict(
  xgb_final,
  dtrain_full,
  predcontrib = TRUE
)

# 5. Plot SHAP
par(oma = c(0, 0, 3, 0)) 
xgb.plot.shap(
  data = dmat_full,
  shap_contrib = shap_contribs,
  model = xgb_final,
  top_n = 2
)

```

